---
title: "ABC Project"
author: "Austin Lesh, Corey Maxedon, Joe Stoica"
date: "12/13/2019"
output: pdf_document
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(plyr)
```

```{r data in, message=FALSE,warning=FALSE}
# read in the data
table_2_1979 <- read_csv("table_2_1979.csv")
table_2_1980 <- read_csv("table_2_1980.csv")
table_3_1975 <- read_csv("table_3_1975.csv")
table_3_1978 <- read_csv("table_3_1978.csv")
```

$q_c$ = the probability that a susceptible individual does not get infected from the community

$q_h$ = the probability that a susceptible individual escapes infection within their household

$w_{js}$ = the probability that j out of the s susceptibles in a household become infected, is given by

We will refer to the following equation as model (1):

$$w_{js} = {s\choose j}w_{jj}(q_cq_h^j)^{s-j}$$

Where 
- $w_{0s} = q_c^s$ for $s=0,1,2,\dots$
- $w_{jj} = 1 - \sum_{i=0}^{j-1}w_{ij}$

```{r prior}
#' Generates four numbers between 0 and 1 and will act as the priors
#'
#' @return a list of four random numbers
simulate_parameters <- function() {
  names = c("qc1", "qh1", "qc2", "qh2")
  
  param = list(runif(1),
               runif(1),
               runif(1),
               runif(1))
  names(param) = names
  return(param)
}
``` 

```{r model 1}
#' The probability that j out of the s susceptibles in a household become infected
#' 
#' @param j number that become infected
#' @param s number of susceptibles
#' @param qc probability that a susceptible individual does not get infected from the community
#' @param qh probability that a susceptible individual does not get infected from their household
#' @param wjj the jth element of the diagonal of w
#' 
#' @return A number between 0 and 1, representing the probability of people infected in a household
model_1 <- function(j, s, qc, qh, wjj){
  return(choose(s, j) * wjj * (qc * qh^j)^(s - j))
}
``` 

```{r generate the data}
#' Generate data from the model specified above
#'
#' @param qc probability that a susceptible individual does not get infected from the community
#' @param qh probability that a susceptible individual does not get infected from their household
#' @param m number of rows of the data
#' @param n number of columns of the data
#'
#' @return a matrix of generated data

data_gen <- function(qc, qh, m, n){
  
  # pre-allocate matrix
  d_star = matrix(NA, nrow = m, ncol = n)
  
  # make first row of matrix using w_{0s} = q_c^s
  d_star[1, ] = qc^(1:n)
  
  # create rest of the data 
  # we will work across the matrix rather than down the columns
  for (j in 2:m) {
    for (s in 1:n) {
      
      # create diagonal data point (when j = j)
      if (j - 1 == s){
        
        # follow the 1 - sum(wij)
        acc = 0      
        for (i in 1:(j-1)) {
          # add the matrix elements above wjj
          acc = acc + d_star[i, j - 1] 
        }
        # populate the matrix with 1 - acc
        d_star[j, j - 1] = 1 - acc
      }
      
      # upper triangular off diagonal calculation
      if (s > j - 1){
        # get the value for wjj for that row
        wjj = d_star[j, j - 1]
        
        # plug it in to the model_1 function to calculate the probability
        d_star[j,s] = model_1(j - 1, s, qc, qh, wjj)
      }
    }
  }
  
  return(d_star)
}
``` 

Use the distance function:

$$d(D_0,D^*)=\frac{1}{2}(||D_1 - D^*(q_{h1},q_{c1})||_F+||D_2 - D^*(q_{h2},q_{c2})||_F)$$

Where
- $|| ||_F$ denotes the Frobenious norm
- $||A||_F=\sqrt{trace(A^TA)}$

- $D_0 = D_1 \cup D_2$, with
- $D_1$ being the 1977-98 outbreak
- $D_2$ being the 1980-81 outbreak
- $D^*$ is the simulation output from model (1) 

```{r frobenious/distance}
#' Calculate the frobenious norm of a matrix
#'
#' @param A the matrix
#'
#' @return the frobenious norm
frobenious <- function(A){
  return(sqrt(sum(abs(A)^2, na.rm = TRUE)))
}
``` 

```{r}
#' Calculating the distance specified above
#'
#' @param d1 The first data matrix (i.e. 1977-78 from table 2)
#' @param d2 The second data matrix (i.e. 1980-81 from table 2)
#' @param d_star1 The first matrix of generated data
#' @param d_star2 The second matrix of generated data
#'
#' @return
distance <- function(d1, d2, d_star1, d_star2){
  # We have to convert the data into proportions from counts
  d1 = as.matrix(d1 / sum(d1, na.rm = TRUE))
  d2 = as.matrix(d2 / sum(d2, na.rm = TRUE))
  # calculate the two frobenious norms
  f1 = frobenious(d1 - d_star1)
  f2 = frobenious(d2 - d_star2)
  
  # this is the distance metric
  ans = (1 / 2) * (f1 + f2) 
  return(ans)
}
```

```{r generate abc function}
#' Generate an ABC sample
#'
#' @param epsilon The specified tolerance levels
#' @param d1 The first data matrix (i.e. 1977-78 from table 2)
#' @param d2 The second data matrix (i.e. 1980-81 from table 2)
#'
#' @return A prior sample that has 
generate_abc_sample <- function(epsilon, d1, d2){
  # drop the first column because we don't need it 
  d1 = d1[, -1]
  d2 = d2[, -1]
  
  # get the dimensions for each dataframe
  dim_d1 = dim(d1)
  dim_d2 = dim(d2)
  
  # basic ABC stuff
  while(TRUE){
    # prior
    prior = simulate_parameters() 
    
    # generate data
    res1 = data_gen(prior$qc1, prior$qh1, m = dim_d1[1], n = dim_d1[2])
    res2 = data_gen(prior$qc2, prior$qh2, m = dim_d2[1], n = dim_d2[2])
    
    # calculate distance
    distance = distance(d1,
                        d2,
                        d_star1 = res1,
                        d_star2 = res2)
    
    # check against tolerance
    if(distance < epsilon){
      break
    }
  }
  return(do.call(rbind, prior))
}
```

```{r results for 3a}
# only uncomment and run if we figure out it's broken, it takes a while
# results = replicate(200, generate_abc_sample(epsilon = 1,
#                                              d1 = table_2_1979,
#                                              d2 = table_2_1980))
# 
# qc1 = results[1,,]
# qh1 = results[2,,]
# qc2 = results[3,,]
# qh2 = results[4,,]
# 
# # this comes from table 2
# res_df <- rbind(data.frame(year = 1977, qh = qh1, qc = qc1), 
#                 data.frame(year = 1980, qh = qh2, qc = qc2))
# write.csv(res_df, "results_3a.csv")
``` 

```{r}
# load in the data that was saved from above
res_df <- read.csv("results_3a.csv")
``` 

```{r}
res_df %>% 
  ggplot(aes(qh, qc, color = factor(year)))+
  geom_point()+
  theme_minimal()+
  scale_x_continuous(breaks = seq(0,1,0.2), limits = c(0,1))+
  scale_y_continuous(breaks = seq(0,1,0.2), limits = c(0,1))+
  labs(color = "Year")
  NULL
```

```{r results for 3c}
#only uncomment and run if we figure out it's broken, it takes a while
# results2 = replicate(200, generate_abc_sample(epsilon = 0.895,
#                                              d1 = table_3_1975,
#                                              d2 = table_3_1978))
# 
# qc12 = results2[1,,]
# qh12 = results2[2,,]
# qc22 = results2[3,,]
# qh22 = results2[4,,]
# 
# # this comes from table 2
# res_df2 <- rbind(data.frame(year = 1975, qh = qh12, qc = qc12),
#                 data.frame(year = 1978, qh = qh22, qc = qc22))
``` 

```{r}
#write.csv(res_df2, "results_3c.csv")

# load in the data that was saved from above
res_df2 <- read.csv("results_3c.csv")
``` 

```{r 3c plot}
res_df2 %>% 
  ggplot(aes(qh, qc, color = factor(year)))+
  geom_point()+
  theme_minimal()+
  scale_x_continuous(breaks = seq(0,1,0.2), limits = c(0,1))+
  scale_y_continuous(breaks = seq(0,1,0.2), limits = c(0,1))+
  labs(color = "Year")+
  NULL
```


# Testing Stuff

```{r tests}
qc = 0.25
qh = 0.75

model_1(j = 1, s = 3, qc = qc, qh =  qh, wjj =  (3/4))
data_gen(qc = 0.25, qh = 0.75, m = 4, n =3)

test1 = data_gen(qc = 0.25, qh = 0.75, m = 4, n =3)

onethree = choose(3,1) * (3/4) * (qc * qh^1)^2
twothree = choose(3,2) * (27/32) * (qc * qh^2)^1
test = matrix(c(0.25, 0.75, NA, NA,
                (1/16), (3/32), (27/32), NA,
                (1/64), onethree, twothree, (1 - onethree - twothree - (1/64))),
         ncol = 3, nrow = 4)

test == test1
``` 

```{r}
#this is to see a good epsilon for 3a
test_f <- function() {
  test = simulate_parameters()
  res1 = data_gen(test$qc1, test$qh1,6,5)
  res2 = data_gen(test$qc2, test$qh2,6,5)

  d = distance(d1 = table_2_1979[, -1],
           d2 = table_2_1980[, -1],
           d_star1 = res1,
           d_star2 = res2)

  return(d)
}

x = replicate(1000, test_f())
quantile(x, seq(0.01, 0.2, 0.01))
``` 


```{r}
#this is to see a good epsilon for 3c
test_f <- function(d1,d2) {
  d1 = d1[, -1]
  d2 = d2[, -1]
  dim_d1 = dim(d1)
  dim_d2 = dim(d2)
  prior = simulate_parameters() 
  
  # generate data
  res1 = data_gen(prior$qc1, prior$qh1, m = dim_d1[1], n = dim_d1[2])
  res2 = data_gen(prior$qc2, prior$qh2, m = dim_d2[1], n = dim_d2[2])
  
  distance = distance(d1,
                      d2,
                      d_star1 = res1,
                      d_star2 = res2)
}

x = replicate(1000, test_f(d1 = table_3_1975, d2 = table_3_1978))
quantile(x, seq(0.01, 0.2, 0.01))
```